### 2017-07-20 chapters/introduction.tex
Latest version. Done with intro to intro. Almost done with physiology intro

Further development in medical research require insight into complex
systems, that is too complicated to study with mathematical models and
super computers.

Further development in medical sciences cannot only rely on Moore's
law but also improvement on mathematical models and numerical
methods.

To lower the risk for the patient we need to more accuratley dioagnose
the patient, and find the optimal treatment.

Personalized medican at the level of the individual.

Many variables to take into account

The story of biomechanics starts in the 17th century with Giovanni
Borelli, known as the father of Biomechanics. He was the first one to
apply the mathematical techniques of Gallileo Galliei to study the
movement of animals.

\input{chapters/introduction/aim_of_thesis}
\input{chapters/introduction/physiology}
% \input{chapters/introduction/mechanics}
\input{chapters/introduction/patient}
\input{chapters/introduction/summary}

\pagebreak

>>>>

% Further development in medical research require insight into complex
% systems, that is too complicated to study with mathematical models and
% super computers.

% Further development in medical sciences cannot only rely on Moore's
% law but also improvement on mathematical models and numerical
% methods.

% To lower the risk for the patient we need to more accuratley dioagnose
% the patient, and find the optimal treatment.

% Personalized medican at the level of the individual.

% Many variables to take into account

% The story of biomechanics starts in the 17th century with Giovanni
% Borelli, known as the father of Biomechanics. He was the first one to
% apply the mathematical techniques of Gallileo Galliei to study the
% movement of animals.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% THE PAST %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \section{Historical note}

One of the earlies examples of modelling applied to cardiac
physiology dates back to 1892 when Robert Wood approximated the
ventricle as a thinned-wall sherical shell.
% In this way, \emph{wall
% stress} could be analytically estimated via the famous \emph{Law of Laplace}
% \cite{woods1892few}. This law relates the wall stress $\sigma$ to the endocardial
% pressure $P$, the endocarial radius $r$ and the wall thickness $w$
% and can in its simplest for be expressed as
% \begin{align}
%   \sigma = \frac{P \times r}{2 w}.
% \end{align}
% Wall stress can be thought of as the load that the heart must work
% against, and it is therefore important to understand how different
% quantities affects the working heart. For instance when the heart
% ejects blood, the load that the heart ejects blood against is
% typically referred to as \emph{afterload}, but this is really the same
% as the wall stress in that particular state. 
% Even though it is obvious that the shape of a ventricle in most cases
% clearly deviates from a sphere, this forumla provides clinician with a tool
% that can be used to understand how diseases affect the heart. Take for
% example the case of ventricular dilation, when the inner radius of 
% the ventricle increases, resulting in increased wall stress and
% hence the heart has to work harder in order to eject blood to the body. In the other
% end we have the case of concentric hypertrophy which results in wall
% thickening and hence reduced the wall stress. In this case more muscle
% fibers are recruited and therefore the tension experienced by each
% individual unit will be less. 
Since 1892 more realistic and sophisticated models have been developed
in order to better describe the anatomical as well as the biophysical
nature of the ventricle. The first extension came in 1927 to
thick-walled spheres \cite{love2013treatise}. In the 1960 the spheres
were substitued by ellipsoids, with thin-walled models in 1963
\cite{sandler1963left}, and by thick-walled models in 1968
\cite{wong1968stress}. A thourough comparison of the different early
models  is presented in \cite{huisman1980comparison}. The first
reported usage of image-based geometries dates back to 1972
\cite{gould1972vivo} were X-ray was used to image the left ventricle
and the finite element method was used to estimate wall stress. The relatively
poor computational power, together with the low image quality really
challenged the biomedical engineers at that time. The following
phrase, taken form \cite{heethaar1977computer} illustrates this fact:  
\emph{``Programming for the solution of the large number of
  simultaneous linear equations generated by the finite element method
  on computers with capacity not exceeding 32K words of memory
  required special attention to very compact storage of the
  stiffness matrix and the retrieval of its coefficients''}.
A paradigm shift in the field of cardiac mechanical modeling came in
the late 1980s at the University of Auckland. Here the first  continuum
description of the heart were developed \cite{hunter1988analysis},
which is still in use today.

% Looking back in time, we see that great progress have been made
% towards personalized medicin. Recent discoveries such as mapping out
% all the genes in the human genome shows that

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% THE PRESENT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Today, advanced imaging techniques allow us to aquire detailed
mechanical information about the heart of the individual. In addition,
advanced numerical algorithms combined with high performance computing
power enable us to get more and more detailed simulation of a patients
heart.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% THE FUTURE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The need for building adequate patient-specific models that captures
the geometrical information observed in the images as well as the
underlying biophysical processes, is recognized as one of the key
challenges in modern bioengnieering \cite{hunter2010vision}.
In the future, models could be used to as a diagnostic tool to extract
features from a patients heart that are otherwise impossible to extract from
images. Eventually, models could potentially predict the outcome of
different treatment strategies and guide clinicians in the decision making.

In other parts of physical sciences and engineering, such as the
space industry or the weather forecast, the role of the computational
science has been cruicial for its development. Without the increasing
computational power predicted by Moore's law
\cite{brock2006understanding} and better numerical methods, such as the
finite element method, scientist would not have been able to predict
complex physical phenomena such as a Tsunami, or being able to send
humans out in space.

Despite our ability to simulate complex physical systems such as
the weather, we still have a long way to go before we are able to
simulate a human organ, such as the heart. The multiscale nature of
biophysical processes has challenged bioengnieers for decades.
Even though we understand processes at each level, such as the flow of
ions in and out of the cell, or the electical transduction through the
heart, the question of how to couple these levels together is still open
\cite{noble2002modeling}.

Another question is how to extract data
observed in medical images and constrain models, in order to make
models specific to the individual. These questions has to be adressed
before we can take the next step into personalized simulations for
\emph{in silico} medicine, and use computer simulations to guide
treatment, and predict outcome.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% THIS THESIS %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Some of these unanswered question serve as motivation for the work
conducted in this thesis. We focus purely on the mechanical modeling
of the heart and the aim has been to develop a framework for
building mechanical models that are personalized at the level of the
individual.

The mechanics of the heart can be described via equations of nonlinear
elasticity which is traditionally formulated using a continuum
mechanics approach. This approach has been extensively used in other
parts of engineering, and numerical solutions to these partial
differential equations (PDE) are solved using the finite element
method (FEM) with the open source software package FEniCS.

These mechanical models contain model parameters which may vary from
individual to individual, and patient specificity can be obtained by
fitting these model parameters to additional clinical data. This
process is often referred to as data assimilation, a field that has
its root in meteorology where one want to make a weather forecast based
on observations from strategically placed weather stations. The idea
is to construct a functional which represent the mismatch between the
model and the observations, and by changing the model parameters we
also change the value of this mismatch functional. Now, what we want to do is to
minimize this functional using these model parameters as control. At
the same time, the governing PDE, which depend upon these model
parameters, has to remain true. We therefore often
refer to such problems as PDE-constrained optimization problems.  
Solving such problems is challenging because evaluation of the
mismatch functional requires the solution of a nonlinear PDE, which is
computationally expensive, and we therefore want to keep the number of
functional evaluations at a minimum. The best way of doing this is
traverse the parameter space along the gradient descent, which is
guaranteed to bring us towards a local minimum. However, this requires
us to compute the functional gradient, or the derivative of the
mismatch functional with respect to the model parameters. As we will
see, the functional gradient can be computed efficiently by
considering the adjoint equation, and recent development in numerical
software tools \cite{farrell2013automated} allow us to exploit this
also numerically.

The remaining introduction in this thesis is organized as follows. In
Section \ref{sec:intro_physiology} we briefly explain some concepts in of
cardiac anatomy and physiology which is used throughout the
thesis. This section ends with a description of the different cardiac
diseases encountered in this thesis. In Section
\ref{sec:intro_mechanical} we give an introduction to continuum
mechanics and mechanical modeling of the heart. The basic equations of
elasticity as well as the constitutive relations for the cardiac
tissue, and some numerical considerations are covered. Finally, in Section 
\ref{sec:intro_personalization} we explain how to personalize a
mechanical model using clinical data. We start by explaining the
different imaging modalities used in this thesis, and how finite
element meshes are created from images. We also explain how to find a
more realistic reference geometry than the one obtained from medical
images. At the end we introduce data assimilation, and how to compute
the functional gradient using the adjoint method.

\newpage
% An even greater challenge

% The problem
% of utilising this information to costrain models to become patient
% specific

% Together with increased computing power, this is perhaps
% the main reason for why computational models today are able to capture
% detailed information

% \begin{itemize}
% \item Building models that are specified at the level of the
%   individual.
% \item Using patient observations to constrain model, while at the same
%   time ensuring that the underlying physical equations are satisfied. 
% \end{itemize}

% \input{chapters/introduction/aim_of_thesis}
\input{chapters/introduction/physiology}
\newpage
\input{chapters/introduction/mechanics}
\newpage
\input{chapters/introduction/patient}
\newpage
\input{chapters/introduction/summary}

\newpage

### 2017-07-20 chapters/introduction/mechanics.tex
Latest version. Done with intro to intro. Almost done with physiology intro

We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$ given by the composition $\varphi = \chi
\circ \chi_0^{-1}$ is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F(\Xvec) = \frac{\partial \varphi}{\partial \Xvec} = \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I
\end{align}
Some other useful quantities are the \emph{right Cauchy-Green} deformation
tensor $\C = \F^T\F$, the \emph{left Cauchy-Green} deformation tensor
$\mathbf{B} = \F\F^T$, the \emph{Green-Lagrange} strain tensor
$\mathbf{E} = \frac{1}{2}(\C - \I)$, and the determinant of the
deformation gradient $J = \det \F$.

@@

\subsection{Incompressibility}
\label{sec:incompressibility}
An incompressible material is a material in which only isochoric
motions are possible. This means that volume of the material do not
change during any deformation and hence we have the constraint
\begin{align}
  J = \det(\F) = 1.
  \label{eq:incompressible_cons}
\end{align}
The constraint \eqref{eq:incompressible_cons} can be imposed by considering the strain energy
function
\begin{align}
  \Psi = \Psi(\F) + p(J-1),
  \label{eq:incomp_strain_energy}
\end{align}
where $p$ is a scalar which serves as a Lagrange multiplier, but which
can be identified as the hydrostatic pressure. Like the displacement,
the hydrostatic pressure is unknown and has to be determined from the equilibrium
equations \eqref{eq:force_balance_strong_ref}. If we differentiate
\eqref{eq:incomp_strain_energy} with respect to $\F$ we get the First
Piola-Kirchhoff stress tensor for an incompressible material
\begin{align}
  \FPK = \frac{\partial \Psi(\F)}{\partial \F} + J p \F^{-T}.
\end{align}
Likewise the Cauchy stress tensor is given by
\begin{align}
  \Cauchy = J^{-1} \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I.
  \label{eq:cauchy_incomp}
\end{align}

@@

Choosing the right boundary conditions for the model is essential,
and the choice should mimic what is observed in the reality. To
physiologically constrain the ventricle in a correct way is difficult,
and many different approaches has been proposed.
The boundary condition and the endocardium is typically modeled as a
Neumann boundary condition, representing the endocardial blood
pressure
\begin{align}
  \sigma \mathbf{n} = -p_{\mathrm{lv}} \mathbf{n}, \;  \xvec \in  \lvendo(t).
\end{align}
This condition has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. 
Note that this condition is imposed on the current configuration, and
to utilize the Lagrangian formulation we can pull back this condition
to the reference configuration to obtain
\begin{align}
  \FPK\mathbf{N} &= -p_{\mathrm{lv}}^{\mathrm{endo}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
\end{align}
Likewise, it is common to enforce a
Neumann boundary condition on the epicardium,
\begin{align}
\FPK\mathbf{N}  &= -p_{\mathrm{lv}}^{\mathrm{epi}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \epi.
\end{align}
However, the pressure $p_{\mathrm{lv}}^{\mathrm{epi}}$ is usually set
to zero. Therefore we drop the superscript and simply referred to the
left ventricular pressure as the left ventricular endocardial
pressure, i.e $p_{\mathrm{lv}} = p_{\mathrm{lv}}^{\mathrm{endo}}$.

There exist a variety of boundary conditions at the base.
It is common to constrain the longitudinal motion of
base, even though it is observed in cardiac images that the apex tend
to be more fixed than the base. A recent study shows that taking into
account the base movement is important to capture the correct
geometrical shape \cite{palit2016passive}. However, this has not been
done in the studies in this thesis.
Fixing the longitudinal motion at the base is enforced through a
Dirichlet boundary condition
\begin{align}
  u_1 = 0,  \;  \Xvec \in \lvbase
\end{align}
To apply this type of condition, it is easiest if the base is
flat and located at a prescribed location, for example in the $x= 0$
plane.
Constraining the longitudinal motion of the base alone is not enough,
since the ventricle is free to move in the basal plane. In order to
anchor the geometry it is possible to fix the movement of the base in
all directions
\begin{align}
  \uvec = \mathbf{0},  \;  \Xvec \in \lvbase,
\end{align}
or fixing the endocardial or epicardial ring
\begin{align}
  \uvec &= \mathbf{0},  \;  \Xvec \in \Gamma_{\mathrm{endo}} \\
  \uvec &= \mathbf{0},  \;  \Xvec \in \Gamma_{\mathrm{epi}}.
\end{align}
Another approach which is used in this thesis is to impose a Robin
type boundary condition at the base
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvbase, 
\end{align}
or at the epicardium to mimic the pericardium
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvepi, 
\end{align}
Here $k$ can be seen as the stiffness of a spring that limit the
movement. The limiting cases, $k = 0$ and $k \rightarrow
\infty$ represent free and fixed boundary respectively.
More complex boundary conditions to mimic the pericardium is also
possible \cite{fritz2014simulation}, but not considered in this thesis.

@@

\begin{remark}
  The basis functions that span the Taylor-Hood
  finite element spaces are also known as the
  Lagrangian basis functions. These basis functions, of
  degree $n$, are a polynomials of degree $n$ on each element, but only
  continuous at the nodes (i.e not continuously
  differentiable). Consequently, differentiating a function that is
  expressed as a linear combination of the Lagrangian basis functions,
  will not be continuous at the nodes, and therefore caution has to
  be made when evaluating features that depends on the derivative of
  such functions. Examples of such features are stress and
  strain with depends on the deformation gradient which again depends
  on the derivative of the displacement. One way to deal with this
  issue is to 1) use other types of elements that are continusly differenetible
  everywhere,  such a the cubic Hermite elements or 2) evaluate the features at the
  Gaussian quadrature points where there is no problem with continuity.
\end{remark}

@@

In this thesis we have in several occasions used a
transversely isotropic version of\eqref{eq:holzapel_full} which is
obtained by setting $a_{fs} = b_{fs}= a_s = b_s = 0$, i.e
\begin{align}
  \label{eq:holzapel_trans}
  \begin{split}
  \Psi(I_1, I_{4\ef}) = \frac{a}{2 b} \left( e^{ b (I_1 - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (I_{4\ef} - 1)_+^2} -1 \right).
  \end{split}
\end{align}
If we further set $a_f = b_f = b = 0$ so that in $a$ is the
only nonzero parameter, then the Holzapfel-Ogden model reduces to (after a
series expansion of the exponential and a limiting argument)
\begin{align}
  \Psi(I_1)  = \frac{a}{2} \left( I_1 - 3 \right), 
\end{align}
which is the model of a Neo Hookean material. The Cauchy stress can be derived
analytically from \eqref{eq:holzapel_full}, by using the chain rule and
\eqref{eq:cauchy_incomp}, 
\begin{align}
  \begin{split}
    \Cauchy
    =& \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I
    = \sum_{i \in \left\{ 1, 4\ef,  4\eS,  8\ef\eS \right\} }
    \psi_i \frac{\partial I_i}{\partial \F}\F^{T} + p \I \\
    =& p \I + a \left( e^{ b (I_1 - 3)}  -1 \right) \mathbf{B} 
    + 2 a_f (I_{4\ef} - 1)_+  e^{ b_f (I_{4\ef} - 1)^2} \mathbf{f} \otimes \mathbf{f} \\
    &+ 2 a_f (I_{4\eS} - 1)_+  e^{ b_f (I_{4\eS} - 1)^2} \mathbf{s} \otimes \mathbf{s} 
    + a_{fs} I_{8\ef\eS}  e^{ b_{fs} I_{8\ef\eS}^2} \left( \mathbf{f} \otimes \mathbf{s}  +  \mathbf{s} \otimes \mathbf{f} \right)
  \end{split}
\end{align}

@@

While the motivation behind the active stress formulation is purely
physiological and based on the Hill model, the motivation behind the
active strain formulation is more driven by ensuring mathematical
robustness. In particular it has been shown \cite{ambrosi2012active}
that with the active strain approach properties such as frame
invariance and rank-one ellipticity is inherited from the strain
energy of the material. In contrast, rank-one elipticity is not
guaranteed for the active stress formulation.

The active stress approach is by far the most used model of active
contraction and is based on the classical Hill model, consisting of a
three element model with one contractile element

Hybrid version \cite{goktepe2014generalized}

@@

\subsubsection{FEniCS}
The FEniCS project started out in 
FEniCS is a framework for solving partial differential
equations using the finite element method. FEniCS has a python
interface where the user can write code which is close to the
mathematical notions used in the theory.

>>>>

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$ given by the composition $\varphi = \chi
\circ \chi_0^{-1}$ is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F(\Xvec) = \frac{\partial \varphi}{\partial \Xvec} = \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I
\end{align}
Some other useful quantities are the \emph{right Cauchy-Green} deformation
tensor $\C = \F^T\F$, the \emph{left Cauchy-Green} deformation tensor
$\mathbf{B} = \F\F^T$, the \emph{Green-Lagrange} strain tensor
$\mathbf{E} = \frac{1}{2}(\C - \I)$, and the determinant of the
deformation gradient $J = \det \F$.

@@

% \subsubsection{Work conjugates}

% \begin{table}[h]
%   \centering
%   \begin{tabular}{ll}
%     \toprule
%     Stress & Strain \\
%     \midrule
%     Second Piola ($\SPiola$) & Green Lagrange ($\mathbf{E}$) \\
%     First Piola ($\FPiola$) & Deformation gradient ($\mathbf{F}$) \\
%     Cauchy ($\mathbf{\sigma}$) &  Eulerian-Almansi ($\mathbf{e}$) \\
%     \bottomrule
%   \end{tabular}
%   \caption{\label{tab:work_conjugates}Work conjugate stress-strain pairs}
% \end{table}

% By conjuagate stress-strain paris we mean that the rate of mechanical
% work is given by the stress times the strain-rate. For example,
% the total work per unit volume done by the stresses from time $t_1$ to
% $t_2$ is given by
% \begin{align}
%   \mathcal{W}(t_1, t_2) &=  \int_{t_1}^{t_2} \mathbf{P} : \dot{\F} dt
%                           = \int_{t_1}^{t_2} \frac{\partial \Psi (\F)}{\partial \mathbf{F}}  : \dot{\F} - p \F^{-T} :  \dot{\F}  dt \\
%                         &= \int_{t_1}^{t_2} \frac{D \Psi (\F)}{D t} - p\cdot 0 dt
%                           = \Psi(\F(t_2)) - \Psi(\F(t_1)).
% \end{align}
% Note that this is a measure of work per unit volume.

\subsection{Incompressibility}
\label{sec:incompressibility}

The myocardium is composed by small blood vessels that supply the
myocardial cells with oxygen. When the myocardium contracts this
perfused blood is then squeezed out, resulting in an overall loss of
2-4\% volume\cite{yin1996compressibility}. Such materials are
referred to as compressible. When the 
volume are preserved we refer to the material as incompressible.
Since 2-4\% is very little, a common assumption in cardiac modeling,
which has also been made in the work conducted in this thesis, is to assume
the myocardium to be incompresible. The reason for this choice is
purely numerical.

For an incompressible material, only isochoric motions are
possible. This means that volume of the material do not change during
any deformation and hence we have the constraint 
\begin{align}
  J = \det(\F) = 1.
  \label{eq:incompressible_cons}
\end{align}
The constraint \eqref{eq:incompressible_cons} can be imposed by
considering the modified strain energy function
\begin{align}
  \Psi = \Psi(\F) + p(J-1),
  \label{eq:incomp_strain_energy}
\end{align}
where $p$ is a scalar which serves as a Lagrange multiplier, but which
can be identified as the hydrostatic pressure. Like the displacement,
the hydrostatic pressure is unknown and has to be determined from the equilibrium
equations \eqref{eq:force_balance_strong_ref}. If we differentiate
\eqref{eq:incomp_strain_energy} with respect to $\F$ we get the First
Piola-Kirchhoff stress tensor for an incompressible material
\begin{align}
  \FPK = \frac{\partial \Psi(\F)}{\partial \F} + J p \F^{-T}.
\end{align}
Likewise the Cauchy stress tensor is given by
\begin{align}
  \Cauchy = J^{-1} \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I.
  \label{eq:cauchy_incomp}
\end{align}

@@

Choosing the correct boundary conditions for the model is essential,
and the choice should mimic what is observed in the reality. To
physiologically constrain the ventricle in a correct way is difficult,
and different approaches has been proposed.
The boundary condition at the endocardium is typically modeled as a
Neumann boundary condition, representing the endocardial blood
pressure, for the left ventricle we have
\begin{align}
  \sigma \mathbf{n} = -p_{\mathrm{lv}} \mathbf{n}, \;  \xvec \in  \lVendo(t), 
\end{align}
and for the right ventricle, ``lv'' is substituted with ``rv''.
This condition has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. 
Note that this condition is imposed on the current configuration, and
to utilize the Lagrangian formulation we can pull back this condition
to the reference configuration to obtain
\begin{align}
  \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lVendo
\end{align}
Likewise, it is common to enforce a
Neumann boundary condition on the epicardium,
\begin{align}
\FPK\mathbf{N}  &= -p_{\mathrm{epi}}  J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \epi.
\end{align}
However, the pressure $p_{\mathrm{lv}}$ is usually set
to zero.

There exist a variety of boundary conditions at the base.
It is common to constrain the longitudinal motion of
base, even though it is observed in cardiac images that the apex tend
to be more fixed than the base. A recent study shows that taking into
account the base movement is important to capture the correct
geometrical shape \cite{palit2016passive}. However, this has not been
done in the studies in this thesis.
Fixing the longitudinal motion at the base is enforced through a
Dirichlet boundary condition,
\begin{align}
  u_1 = 0,  \;  \Xvec \in \lvbase,
\end{align}
where $u_1$ is the longitudinal component of the displacement $\uvec =
(u_1, u_2, u_3)$. To apply this type of condition, it is easiest if
the base is flat and located at a prescribed location, for example in
the $x= 0$ plane. Constraining the longitudinal motion of the base
alone is not enough since the ventricle is free to move in the basal
plane. In order to anchor the geometry it is possible to fix the
movement of the base in all directions
\begin{align}
  \uvec = \mathbf{0},  \;  \Xvec \in \lvbase,
\end{align}
or fixing the endocardial or epicardial ring
\begin{align}
  \uvec &= \mathbf{0},  \;  \Xvec \in \Gamma_{\mathrm{endo}} \\
  \uvec &= \mathbf{0},  \;  \Xvec \in \Gamma_{\mathrm{epi}}.
\end{align}

Another approach which is used in this thesis is to impose a Robin
type boundary condition at the base
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvbase, 
\end{align}
or at the epicardium to mimic the pericardium
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvepi, 
\end{align}
Here $k$ can be seen as the stiffness of a spring that limit the
movement. The limiting cases, $k = 0$ and $k \rightarrow
\infty$ represent free and fixed boundary respectively.
More complex boundary conditions to mimic the pericardium is also
possible \cite{fritz2014simulation}, but not considered in this thesis.
An overview of the location of the different boundaries for the
bi-ventricular geometry is illustrated in Figure \ref{fig:boundaries}.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/boundaries}
\caption{Illustration of the different boundaries in a bi-ventricular
  domain.}
\label{fig:boundaries}
\end{figure}

@@

\begin{remark}
  The basis functions that span the Taylor-Hood
  finite element spaces are also known as the
  Lagrangian basis functions. These basis functions, of
  degree $n$, are a polynomials of degree $n$ on each element, but only
  continuous at the nodes (i.e not continuously
  differentiable). Consequently, differentiating a function that is
  expressed as a linear combination of the Lagrangian basis functions,
  will not be continuous at the nodes, and therefore caution has to
  be made when evaluating features that depends on the derivative of
  such functions. Examples of such features are stress and
  strain with depends on the deformation gradient which again depends
  on the derivative of the displacement. One way to deal with this
  issue is to 1) use other types of elements that are continusly
  differentiable everywhere,  such a the cubic Hermite elements or 2) evaluate the features at the
  Gaussian quadrature points where there is no problem with continuity.
\end{remark}

@@

In this thesis we have in several occasions used a
transversely isotropic version of \eqref{eq:holzapel_full} which is
obtained by setting $a_{fs} = b_{fs}= a_s = b_s = 0$, i.e
\begin{align}
  \label{eq:holzapel_trans}
  \begin{split}
  \Psi(I_1, I_{4\ef}) = \frac{a}{2 b} \left( e^{ b (I_1 - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (I_{4\ef} - 1)_+^2} -1 \right).
  \end{split}
\end{align}
If we further set $a_f = b_f = b = 0$ so that in $a$ is the
only nonzero parameter, then the Holzapfel-Ogden model reduces to (after a
series expansion of the exponential and a limiting argument)
\begin{align}
  \Psi(I_1)  = \frac{a}{2} \left( I_1 - 3 \right), 
\end{align}
which is the model of a Neo Hookean material. The Cauchy stress can be derived
analytically from \eqref{eq:holzapel_full}, by using the chain rule and
\eqref{eq:cauchy_incomp}, 
\begin{align}
  \begin{split}
    \Cauchy
    =& \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I
    = \sum_{i \in \left\{ 1, 4\ef,  4\eS,  8\ef\eS \right\} }
    \psi_i \frac{\partial I_i}{\partial \F}\F^{T} + p \I \\
    =& p \I + a \left( e^{ b (I_1 - 3)}  -1 \right) \mathbf{B} 
    + 2 a_f (I_{4\ef} - 1)_+  e^{ b_f (I_{4\ef} - 1)^2} \mathbf{f} \otimes \mathbf{f} \\
    &+ 2 a_f (I_{4\eS} - 1)_+  e^{ b_f (I_{4\eS} - 1)^2} \mathbf{s} \otimes \mathbf{s} 
    + a_{fs} I_{8\ef\eS}  e^{ b_{fs} I_{8\ef\eS}^2} \left( \mathbf{f} \otimes \mathbf{s}  +  \mathbf{s} \otimes \mathbf{f} \right),
  \end{split}
\end{align}
where $\mathbf{B} = \F \F^{T}$ is the left Cauchy-Green tensor,
$\mathbf{f} = \F \ef$ and $\mathbf{s} = \F \eS$.

@@

While the motivation behind the active stress formulation is purely
physiological and based on the classical Hill model shown in Figure
\ref{fig:hill_muscle_model}, the motivation behind the active strain
formulation is more driven by ensuring mathematical robustness. In
particular it has been shown \cite{ambrosi2012active} that with the
active strain formaulation properties such as frame invariance and
rank-one ellipticity is inherited from the strain energy function. In
contrast, rank-one elipticity is not guaranteed for the active stress
formulation.

% Other types of active models also exists \cite{goktepe2014generalized}
% but will not be discssed f

@@

\subsubsection{FEniCS}
\label{sec:fenics}
The FEniCS project started out in 
FEniCS is a framework for solving partial differential
equations using the finite element method. FEniCS has a python
interface where the user can write code which is close to the
mathematical notions used in the theory.

### 2017-07-20 chapters/introduction/patient.tex
Latest version. Done with intro to intro. Almost done with physiology intro

\section{Personalization of Cardiac Biomechanics}

In this section we will go through the steps needed to personalize the
mechanical models developed in the previous section towards a given
patient. We will start by giving a brief overview of the
interaction between medical imaging and cardiac computational model.
Then we will explain how the geometries are extracted from the
medical images and turned into a computational mesh,  which we can use
for finite element computations. Finally we will explain how to
incorporate additional data into model by means of adjoint based data
assimilation techniques.

>>>>

\section{Personalization of Cardiac Biomechanics}
\label{sec:intro_personalization}
In this section we will go through the steps needed to personalize the
mechanical models developed in the previous section towards a given
patient. We will start by giving a brief overview of the
interaction between medical imaging and cardiac computational model.
Then we will explain how the geometries are extracted from the
medical images and turned into a computational mesh,  which we can use
for finite element computations. Finally we will explain how to
incorporate additional data into model by means of adjoint based data
assimilation techniques.
